### 2018-03-07 chapter4/ms.tex
Chapter 4 First Pass

\section{Introduction}
\label{sec:intro}

The Advanced Laser Interferometer Gravitational-Wave Observatory
(LIGO) is designed to detect the final inspiral and
merger of compact object binaries comprised of neutron stars (NS)
and/or stellar-mass black holes (BH) \citep{abb+09}. 
The first LIGO observing run (designated O1)
commenced on 18 September 2015 with the ability to detect binary
neutron star (BNS) mergers to an average distance of $\sim 80$ Mpc, a
forty-fold increase in volume relative to the previous generation of
ground-based GW detectors. On 2015 September 14 LIGO detected 
the first GW event ever observed, GW150914 \citep{abb+16a}.

The waveform of GW150914 was consistent with the inspiral, merger,
and ring-down of a binary black hole (BBH) system ($36+29$ M$_\odot$; 
\citealt{abb+16a}) providing the first observational
evidence that such systems exist and merge. While there are no robust
theoretical predictions for the expected electromagnetic (EM)
counterparts of such a merger, more than 20 teams conducted a wide
range of follow-up observations spanning from radio to $\gamma$-rays, along with neutrino follow up
\citep{abb+16b, adrian+16, annis+16, conn+16, evans+16, kasliwal+16, sav+16, smartt+16, ss+16, tavani+16}.
This effort included deep optical follow-up observations by our group using DECam covering 100 deg$^2$
(corresponding to 38\%~(11\%) of the initial~(final) sky maps)
 -- making this one of the most comprehensive optical 
follow-up campaigns for GW150914 \citep{ss+16,annis+16}. Our search for rapidly
declining transients to limiting magnitudes of $i\approx21.5$ mag
for red $(i-z=1)$ and $i\approx20.1$ mag for blue $(i-z=-1)$ events yielded no counterpart to
GW150914 \citep{ss+16}. One result of the broader multi-wavelength
follow-up campaign is a claimed coincident detection of a weak short
gamma-ray burst (SGRB) from the {\it Fermi}/GBM detector 0.4 s after
the GW event \citep{conn+16}. However, this event was not
detected in INTEGRAL $\gamma$-ray data \citep{sav+16} and was also
disputed in a re-analysis of the GBM data \citep{greiner+16}.

A second high-significance GW event, designated GW151226, was
discovered by LIGO on 2015 December 26 at 03:38:53 UT \citep{abb+16c}. 
This event was also due to the inspiral and merger of a BBH system, consisting 
of $14.2^{+8.3}_{-3.7}$~M$_\odot$ and $7.5^{+2.3}_{-2.3}$~M$_\odot$ black holes at a luminosity 
distance of $d_L = 440^{+180}_{-190}$~Mpc \citep{abb+16c}.
The initial localization was provided as a probability sky map via a private 
GCN circular 37 hours after the GW detection \citep{sing+15}. We initiated
optical follow-up observations with DECam 10 hours later on 2015 December 28,
and imaged a 28.8 deg$^2$ region in the $i$ and $z$ bands during several epochs. Here we report 
the results of this search. In Section~\ref{sec:obs} we discuss the
observations and data analysis procedures. In Section~\ref{sec:analysis} 
we present our search methodology for potential counterparts to GW151226, 
and the results of this search. We summarize our conclusions in Section~\ref{sec:conc}. We perform 
cosmological calculations assuming $H_0 = 67.8$~km  s$^{-1}$ Mpc$^{-1}$,
$\Omega_{\lambda} = 0.69$, and $\Omega_m = 0.31$ \citep{planck15}. 
 Magnitudes are reported in the AB system.

\section{Observations and Data Reduction} 
\label{sec:obs}

GW151226 was detected on 2015 December 26 at 03:38:53 UT by a 
Compact Binary Coalescence (CBC) search pipeline \citep{abb+16c}. The CBC
pipeline operates by matching the strain data against waveform
templates and is sensitive to mergers containing NS and/or BH. 
The initial sky map was generated by the {\tt BAYESTAR} algorithm
and released 37 hours after the GW detection. {\tt BAYESTAR} is a Bayesian 
algorithm that generates a localization sky map based on the parameter
estimation from the CBC pipeline \citep{sing+14,sing+16}. The sky area contained within the
initial 50\% and 90\% contours was 430 deg$^2$ and 1340 deg$^2$,
respectively. A sky map generated by the {\tt LALInference}
algorithm, which computes the localization using Bayesian forward-modeling 
of the signal morphology \citep{veitch+15}, was released on 2016 January 15 UT, 
after our DECam observations had been concluded. The {\tt LALInference} sky map
is slightly narrower than the sky map from {\tt BAYESTAR} with 50\% and 90\% contours 
of 362 deg$^2$ and 1238 deg$^2$, respectively.

\begin{figure}[t!]
\centering
\includegraphics[width=0.75 \textwidth]{./figs/chapter4/fig1.pdf}  
\caption{
Sky region covered by our DECam observations (red
hexagons) relative to the 50\% and 90\% probability regions 
from the {\tt BAYESTAR} (cyan contours) and {\tt LALInference}
(white contours) localization of GW151226. 
The background color indicates the estimated $5\sigma$ 
point-source limiting magnitude for a 90~s $i$-band exposure 
as a function of sky position for the first night of our DECam 
observations. The variation in the limiting magnitude is largely 
driven by the dust extinction and airmass at that position. 
The dark grey regions indicate sky positions that 
were unobservable due to the telescope pointing limits.  The yellow contour 
indicates the region of sky covered by the Dark Energy Survey 
(DES). The total effective area for the 12 DECam pointings is 28.8 
deg$^2$, corresponding to 3\%~(2\%) of the probability in the
{\tt BAYESTAR}~({\tt LALInference}) sky map.}
\label{fig:obs}
\end{figure}

We initiated follow-up observations with DECam on 2015 December 28 UT,
two days after the GW detection and 10 hours after distribution of the {\tt BAYESTAR}
sky map. DECam is a wide-field optical imager with a 3 deg$^2$ field of view 
\citep{flaugher+15}. We imaged a 28.8 deg$^2$ region corresponding to 3\%
of the sky localization probability when convolved with the initial {\tt BAYESTAR} map 
and 2\% of the localization probability in the final {\tt LALInference} sky map. 
The pointings and ordering of the DECam observations were determined using
the automated algorithm described in Soares-Santos et al. \citeyearpar{ss+16}. The choice 
of observing fields was constrained by weather, instrument availability, and the available
time to observe this sky region given its high airmass. We
obtained four epochs of data with each epoch consisting of one 90 s
exposure in $i$-band and two 90 s exposures in $z$-band for each of the 12 pointings. The first
epoch was obtained 2--3 days after the initial LIGO trigger (2015
December 28--29 UT), the second epoch was at 6 days (2016 January 1
UT), the third epoch was at 13--14 days (2016 January 8--9), and the
fourth epoch was at 23--24 days (2016 January 18--19). A summary of
the observations is provided in Table~\ref{datatable} and a visual
representation of the sky region is shown in
Figure~\ref{fig:obs}.

We processed the data using an implementation of the {\tt photpipe} 
pipeline modified for DECam images. {\tt Photpipe} is a
pipeline used in several time-domain surveys (e.g.,
SuperMACHO, ESSENCE, Pan-STARRS1; see \citealt{rest+05,garg+07,mik+07,rest+14}), designed to perform
single-epoch image processing including image calibration (e.g., bias
subtraction, cross-talk corrections, flat-fielding), astrometric
calibration, image coaddition, and photometric calibration.
Additionally, {\tt photpipe} performs difference imaging using {\tt
  hotpants} \citep{alard00,becker15} to compute a spatially varying convolution kernel,
followed by photometry on the difference images using an
implementation of {\tt DoPhot} optimized for point spread function (PSF) photometry on
difference images \citep{schechter+93}. Lastly, we
use {\tt photpipe} to perform initial candidate searches by
specifying a required number of spatially coincident detections over a
range of time. Once candidates are identified, {\tt photpipe} performs
``forced" PSF photometry on the subtracted images at the fixed coordinates of
an identified candidate in each available epoch.

In the case of the GW151226 observations, we began with raw images
acquired from the NOAO archive\footnote{http://archive.noao.edu/} and the
most recent calibration files\footnote{http://www.ctio.noao.edu/noao/content/decam-calibration-files}.
Astrometric calibration was performed relative to the Pan-STARRS1 (PS1)
$3\pi$ survey and 2MASS $J$-band catalogs. The two $z$-band exposures
were then coadded. Photometric calibration was performed using the
PS1 $3\pi$ survey with appropriate calibrations between 
PS1 and DECam magnitudes \citep{scolnic+15}. Image subtraction was performed
using observations from the final epoch as templates. The approach
to candidate selection is described in Section~\ref{sec:analysis}.

\begin{deluxetable}{lrccccccc}
\tabletypesize{\footnotesize}
\tablecolumns{9}
\tablewidth{0pt}
\tablecaption{Summary of DECam Observations of GW151226
\label{datatable}}
\tablehead{	
\colhead{Visit}    &  
\colhead{UT}       & 
\colhead{$\Delta t$\tablenotemark{a}} &
\colhead{$\langle$PSF$_i$$\rangle$}    &
\colhead{$\langle$PSF$_z$$\rangle$}    &
\colhead{$\langle$airmass$\rangle$}  &
\colhead{$\langle$depth$_i\rangle$}   &
\colhead{$\langle$depth$_z\rangle$}  &
\colhead{$A_{\mathrm{eff}}$\tablenotemark{b}} \\
\colhead{ } & 
\colhead{ } &
\colhead{(days)} &
\colhead{(arcsec)} &
\colhead{(arcsec)} &
\colhead{ } &
\colhead{(mag)} &
\colhead{(mag)} &
\colhead{(deg$^2$)} 
}
\startdata
 Epoch 1 & 2015-12-28.11 & 1.96 & 0.97 & 0.99 & 1.95 & 22.39 & 22.23 & 14.4 \\      %  6 hexes 
               & 2015-12-29.11 & 2.96 & 1.00 & 0.97 & 1.78 & 22.57 & 22.46 & 14.4 \\      %  6 hexes
                \hline
 Epoch 2 & 2016-01-01.06 & 5.91 & 0.95 & 0.90 & 1.57 & 21.37 & 21.06 & 28.8 \\      % 12 hexes
 \hline
 Epoch 3 & 2016-01-08.11 & 12.96 & 1.68 & 1.62 & 2.15 & 22.09 & 21.70 & 24.0 \\      % 10 hexes
                & 2016-01-09.11 & 13.96 & 1.17 & 1.12 & 1.80 & 22.44 & 22.17 &  4.8 \\      %  2 hexes
                 \hline
 Epoch 4 & 2016-01-18.03 & 22.88 & 1.21 & 1.20 & 1.48 & 22.00 & 22.01 & 12.0 \\      %  5 hexes
               & 2016-01-19.01 & 23.86 & 1.29 & 1.25 & 1.71 & 21.86 & 21.90 & 16.8 \\ %  7 hexes
\enddata
\tablecomments{Summary of our DECam follow-up observations of
GW151226. The PSF, airmass, and depth are the average values across
all observations on that date. The reported depth corresponds to the mean 
$5\sigma$ point source detection in the coadded search images.}
\tablenotetext{a}{Time elapsed between the GW trigger time and the time of the first image.} 
\tablenotetext{b}{The effective area corresponds to 12 DECam pointings 
taking into account that $\approx 20\%$ of the 3
deg$^2$ field of view of DECam is lost due to chip gaps (10\%), 3
dead CCDs (5\%, \citealt{diehl+14}), and masked edge pixels (5\%).}
\end{deluxetable}

@@

\section{Search for an Optical Counterpart}
\label{sec:analysis}

The primary focus of our search is a fast-fading transient. 
While the merger of a BBH system is not expected
to produce an EM counterpart, it is informative to consider the
possibility of optical emission due to the presence of some matter in
the system. As a generic example, we consider the behavior of a
transient such as a short gamma-ray burst (SGRB) with a typical beaming-corrected energy of
$E_j \approx 10^{49}$ erg and an opening angle of $\theta_j\approx
10^\circ$ \citep{berger14,fong+15}. If viewed far off-axis $(\theta_{\rm
  obs}\gtrsim 4\theta_j)$ the optical emission will reach peak
brightness after several days, but at the distance of GW151226
($\approx440$ Mpc, \citealt{abb+16c}), the peak brightness will be $i\approx 26$ mag (see Figure 5 of
\citealt{metzger12}), well beyond our detection limit. If the source
is observed moderately off-axis or on-axis $(\theta_{\rm obs} \lesssim
2\theta_j)$, then the light curve will decline throughout our
observations, roughly as $F_\nu\propto t^{-1}$, and will be detectable
at $i\approx$ 21--22 mag in our first observation (see Figures 3 and 4
of \citealt{metzger12}). We can apply a similar argument to the
behavior of a more isotropic (and non-relativistic) outflow given that
any material ejected in a BBH merger is likely to have a low mass and
the outflow will thus become optically thin early, leading to fading
optical emission. Based on this line of reasoning, we search our data
for steadily declining transients.

@@

Only four events passed our final criterion. We find that two of those events are 
coincident with the nuclei of known AGN (PKS\,0129-066 and Mrk\,584), indicating 
that they represent AGN variability. A third candidate is coincident with the nucleus of
the bright radio source PMN\,J0203+0956 ($F_\nu(365\,{\rm MHz})\approx 0.4$ Jy, 
\citealt{douglas+96}), also suggesting AGN variability.

\begin{figure}[t!]
\centering
\includegraphics[width=0.95\textwidth]{./figs/chapter4/fig2.pdf}
\caption{{\it Top}: Single-epoch images of our main candidate
from all four epochs (green circle). This is the event discovered as PS15cdi in
the PSST about 94 d prior to GW151226. {\it Bottom}: Light
curve data for PS15cdi from PSST $w$- and $i$-band
observations (green squares and yellow diamonds, respectively).
Our DECam $i$- and $z$-band data are shown as
blue circles and red stars, respectively. The revised DECam analysis 
using pre-existing templates is shown as open symbols. Upper
limits are indicated by triangles. The inset focuses on our
DECam data, indicating a rapid decline in both $i$ and $z$
bands.  We fit a power-law model to the data finding a temporal index of 
$\alpha = -0.43$ (dashed-dot line). Kilonova models from Barnes and 
Kasen \citeyearpar{barnes13} with $v_{\rm ej}=0.2c$ and $M_{\rm ej}=0.1$ M$_\odot$ 
at a distance of $187$~Mpc are also shown (dashed line).}
\label{fig:PS15cdi}
\end{figure}

The final candidate in our search is located at RA = 01$^{\rm h}$42$^{\rm m}$16$^{\rm s}$.17 
and DEC = $-$02\arcdeg13\arcmin42 6\arcsec~(J2000), with an offset of $5.8$ arcsec from the galaxy 
CGCG 386-030 (RA =  01$^{\rm h}$42$^{\rm m}$15$^{\rm s}$.6, DEC = $-$02\arcdeg13\arcmin38 
5\arcsec; J2000), at $z = 0.041$ or $d_L \approx 187$ Mpc (6dFGS, \citealt{jones+04,jones+09}); see 
Figure~\ref{fig:PS15cdi}. We note that this distance is inconsistent with the 90\% confidence interval for
the distance to GW151226 based on the GW data \citep{abb+16c}. We observe this source in a state of rapid decline with an 
absolute magnitude of $M_i \approx -15$~mag on 2015 December 28 and $M_i\approx -14.5$~mag on 2016 
January 1, indicating a decline rate of $\approx 0.12$ mag d$^{-1}$; the decline rate in $z$-band is 
$\approx 0.10$ mag d$^{-1}$. Additionally, the source exhibits a red $i-z$ color of $0.3$ mag. We 
fit these data to a power-law model typical for GRB afterglows $(F_\nu \propto \nu^{\beta} t^{\alpha})$ 
and find a temporal index of $\alpha = -0.43\pm0.12$ and a spectral index of $\beta = -1.8\pm0.8$, 
both of which differ from the expected values for GRB afterglows ($\alpha \approx -1$, 
$\beta \approx -0.75$, \citealt{sari+98}). Additionally, we compare our observations to a 
kilonova model with ejecta parameters of $v_{\rm ej} = 0.2$c and $M_{\rm ej} = 0.1$ M$_\odot$
 \citep{barnes13}. We find that the timescale of the transient agrees with those expected for kilonovae, 
 but the color is bluer than the expected value of $i-z \approx 1$~mag \citep{barnes13}. Thus, the properties
 of this transient differ from those of GRB afterglows or kilonovae. The observations 
 and models are shown in Figure~\ref{fig:PS15cdi}.

This source was previously detected as PS15cdi on 2015 September 23 by the Pan-STARRS 
Survey for Transients (PSST\footnote{\url{http://psweb.mp.qub.ac.uk/ps1threepi/psdb/candidate/1014216170021342600/}},
\citealt{huber+15}); see Figure~\ref{fig:PS15cdi}. The absolute $i$-band 
magnitude in the first PSST epoch, $M_i\approx -16.6$ mag and the shallow decline of $\approx0.6$~mag 
over $\approx70$~d, are consistent with a Type IIP core-collapse supernova (SN). 
A likely interpretation of the rapid decline in our observations is that PS15cdi is a Type IIP SN undergoing 
the rapid transition from the hydrogen recombination driven plateau to the radioactive $^{56}$Co dominated 
phase \citep{kasen09,dhun+16,sanders+15}. The red $i-z$ color in our data is consistent with observations 
of other IIP SN during this phase of evolution (e.g., SN2013ej, \citealt{dhun+16}).
This transition typically occurs about 100~d post explosion \citep{kasen09,dhun+16,sanders+15}, consistent with the 
timing of our observations relative to the first detection in PSST.

>>>>

\section{Introduction}
\label{sec:ch4_intro}

The Advanced Laser Interferometer Gravitational-Wave Observatory
(LIGO) is designed to detect the final inspiral and
merger of compact object binaries comprised of neutron stars (NS)
and/or stellar-mass black holes (BH) \citep{LIGOMainRef}.
The first LIGO observing run (designated O1)
commenced on 18 September 2015 with the ability to detect binary
neutron star (BNS) mergers to an average distance of $\sim 80$ Mpc, a
forty-fold increase in volume relative to the previous generation of
ground-based GW detectors. On 2015 September 14 LIGO detected
the first GW event ever observed, GW150914 \citep{LIGOGW150914}.

The waveform of GW150914 was consistent with the inspiral, merger,
and ring-down of a binary black hole (BBH) system ($36+29$ M$_\odot$;
\citealt{LIGOGW150914}) providing the first observational
evidence that such systems exist and merge. While there are no robust
theoretical predictions for the expected electromagnetic (EM)
counterparts of such a merger, more than 20 teams conducted a wide
range of follow-up observations spanning from radio to $\gamma$-rays, along with neutrino follow up
\citep{LIGOGW150914FollowUp,LIGOGW150914FollowUpSupp,GW150914IceCube,Annis+16,GW150914Fermi,Evans+16,Kasliwal+16,Savchenko+16,GW150914PS1,GW150914DECam,Tavani+16}.
This effort included deep optical follow-up observations by our group using DECam covering 100 deg$^2$
(corresponding to 38\%~(11\%) of the initial~(final) sky maps)
 -- making this one of the most comprehensive optical
follow-up campaigns for GW150914 \citep{Annis+16,GW150914DECam}. Our search for rapidly
declining transients to limiting magnitudes of $i\approx21.5$ mag
for red $(i-z=1)$ and $i\approx20.1$ mag for blue $(i-z=-1)$ events yielded no counterpart to
GW150914 \citep{GW150914DECam}. One result of the broader multi-wavelength
follow-up campaign is a claimed coincident detection of a weak short
gamma-ray burst (SGRB) from the {\it Fermi}/GBM detector 0.4 s after
the GW event \citep{GW150914Fermi}. However, this event was not
detected in INTEGRAL $\gamma$-ray data \citep{Savchenko+16} and was also
disputed in a re-analysis of the GBM data \citep{Greiner+16}.

A second high-significance GW event, designated GW151226, was
discovered by LIGO on 2015 December 26 at 03:38:53 UT \citep{abb+16c}.
This event was also due to the inspiral and merger of a BBH system, consisting
of $14.2^{+8.3}_{-3.7}$~M$_\odot$ and $7.5^{+2.3}_{-2.3}$~M$_\odot$ black holes at a luminosity
distance of $d_L = 440^{+180}_{-190}$~Mpc \citep{LIGOGW151226}.
The initial localization was provided as a probability sky map via a private
GCN circular 37 hours after the GW detection \citep{SingerPrice16}. We initiated
optical follow-up observations with DECam 10 hours later on 2015 December 28,
and imaged a 28.8 deg$^2$ region in the $i$ and $z$ bands during several epochs. Here we report
the results of this search. In \cref{sec:ch4_obs} we discuss the
observations and data analysis procedures. In \cref{sec:ch4_analysis}
we present our search methodology for potential counterparts to GW151226,
and the results of this search. We summarize our conclusions in \cref{sec:ch4_conc}. We perform
cosmological calculations assuming $H_0 = 67.8$~km  s$^{-1}$ Mpc$^{-1}$,
$\Omega_{\lambda} = 0.69$, and $\Omega_m = 0.31$ \citep{Planck2016}.
 Magnitudes are reported in the AB system.

\section{Observations and Data Reduction}
\label{sec:ch4_obs}

GW151226 was detected on 2015 December 26 at 03:38:53 UT by a
Compact Binary Coalescence (CBC) search pipeline \citep{LIGOGW151226}. The CBC
pipeline operates by matching the strain data against waveform
templates and is sensitive to mergers containing NS and/or BH.
The initial sky map was generated by the {\tt BAYESTAR} algorithm
and released 37 hours after the GW detection. {\tt BAYESTAR} is a Bayesian
algorithm that generates a localization sky map based on the parameter
estimation from the CBC pipeline \citep{Singer+14,SingerPrice16}. The sky area contained within the
initial 50\% and 90\% contours was 430 deg$^2$ and 1340 deg$^2$,
respectively. A sky map generated by the {\tt LALInference}
algorithm, which computes the localization using Bayesian forward-modeling
of the signal morphology \citep{Veitch+15}, was released on 2016 January 15 UT,
after our DECam observations had been concluded. The {\tt LALInference} sky map
is slightly narrower than the sky map from {\tt BAYESTAR} with 50\% and 90\% contours
of 362 deg$^2$ and 1238 deg$^2$, respectively.

\begin{figure}[t!]
\centering
\includegraphics[width=0.75 \textwidth]{./figs/chapter4/fig1.pdf}
\caption{
Sky region covered by our DECam observations (red
hexagons) relative to the 50\% and 90\% probability regions
from the {\tt BAYESTAR} (cyan contours) and {\tt LALInference}
(white contours) localization of GW151226.
The background color indicates the estimated $5\sigma$
point-source limiting magnitude for a 90~s $i$-band exposure
as a function of sky position for the first night of our DECam
observations. The variation in the limiting magnitude is largely
driven by the dust extinction and airmass at that position.
The dark grey regions indicate sky positions that
were unobservable due to the telescope pointing limits.  The yellow contour
indicates the region of sky covered by the Dark Energy Survey
(DES). The total effective area for the 12 DECam pointings is 28.8
deg$^2$, corresponding to 3\%~(2\%) of the probability in the
{\tt BAYESTAR}~({\tt LALInference}) sky map.}
\label{fig:ch4_obs}
\end{figure}

We initiated follow-up observations with DECam on 2015 December 28 UT,
two days after the GW detection and 10 hours after distribution of the {\tt BAYESTAR}
sky map. DECam is a wide-field optical imager with a 3 deg$^2$ field of view
\citep{Flaugher+15}. We imaged a 28.8 deg$^2$ region corresponding to 3\%
of the sky localization probability when convolved with the initial {\tt BAYESTAR} map
and 2\% of the localization probability in the final {\tt LALInference} sky map.
The pointings and ordering of the DECam observations were determined using
the automated algorithm described in \citet{GW150914DECam}. The choice
of observing fields was constrained by weather, instrument availability, and the available
time to observe this sky region given its high airmass. We
obtained four epochs of data with each epoch consisting of one 90 s
exposure in $i$-band and two 90 s exposures in $z$-band for each of the 12 pointings. The first
epoch was obtained 2--3 days after the initial LIGO trigger (2015
December 28--29 UT), the second epoch was at 6 days (2016 January 1
UT), the third epoch was at 13--14 days (2016 January 8--9), and the
fourth epoch was at 23--24 days (2016 January 18--19). A summary of
the observations is provided in \cref{datatable} and a visual
representation of the sky region is shown in
\Cref{fig:ch4_obs}.

We processed the data using an implementation of the {\tt photpipe}
pipeline modified for DECam images. {\tt Photpipe} is a
pipeline used in several time-domain surveys (e.g.,
SuperMACHO, ESSENCE, Pan-STARRS1; see \citealt{Rest+05,Garg+07,Miknaitis+07,Rest+14}), designed to perform
single-epoch image processing including image calibration (e.g., bias
subtraction, cross-talk corrections, flat-fielding), astrometric
calibration, image coaddition, and photometric calibration.
Additionally, {\tt photpipe} performs difference imaging using {\tt
  hotpants} \citep{Alard2000,Becker2015} to compute a spatially varying convolution kernel,
followed by photometry on the difference images using an
implementation of {\tt DoPhot} optimized for point spread function (PSF) photometry on
difference images \citep{Schechter+93}. Lastly, we
use {\tt photpipe} to perform initial candidate searches by
specifying a required number of spatially coincident detections over a
range of time. Once candidates are identified, {\tt photpipe} performs
``forced" PSF photometry on the subtracted images at the fixed coordinates of
an identified candidate in each available epoch.

In the case of the GW151226 observations, we began with raw images
acquired from the NOAO archive\footnote{http://archive.noao.edu/} and the
most recent calibration files\footnote{http://www.ctio.noao.edu/noao/content/decam-calibration-files}.
Astrometric calibration was performed relative to the Pan-STARRS1 (PS1)
$3\pi$ survey and 2MASS $J$-band catalogs. The two $z$-band exposures
were then coadded. Photometric calibration was performed using the
PS1 $3\pi$ survey with appropriate calibrations between
PS1 and DECam magnitudes \citep{Scolnic+15}. Image subtraction was performed
using observations from the final epoch as templates. The approach
to candidate selection is described in \cref{sec:ch4_analysis}.

\begin{deluxetable}{lrccccccc}
\tabletypesize{\footnotesize}
\tablecolumns{9}
\tablewidth{0pt}
\tablecaption{Summary of DECam Observations of GW151226
\label{datatable}}
\tablehead{
\colhead{Visit}    &
\colhead{UT}       &
\colhead{$\Delta t$\tablenotemark{a}} &
\colhead{$\langle$PSF$_i$$\rangle$}    &
\colhead{$\langle$PSF$_z$$\rangle$}    &
\colhead{$\langle$airmass$\rangle$}  &
\colhead{$\langle$depth$_i\rangle$}   &
\colhead{$\langle$depth$_z\rangle$}  &
\colhead{$A_{\mathrm{eff}}$\tablenotemark{b}} \\
\colhead{ } &
\colhead{ } &
\colhead{(days)} &
\colhead{(arcsec)} &
\colhead{(arcsec)} &
\colhead{ } &
\colhead{(mag)} &
\colhead{(mag)} &
\colhead{(deg$^2$)}
}
\startdata
 Epoch 1 & 2015-12-28.11 & 1.96 & 0.97 & 0.99 & 1.95 & 22.39 & 22.23 & 14.4 \\      %  6 hexes
               & 2015-12-29.11 & 2.96 & 1.00 & 0.97 & 1.78 & 22.57 & 22.46 & 14.4 \\      %  6 hexes
                \hline
 Epoch 2 & 2016-01-01.06 & 5.91 & 0.95 & 0.90 & 1.57 & 21.37 & 21.06 & 28.8 \\      % 12 hexes
 \hline
 Epoch 3 & 2016-01-08.11 & 12.96 & 1.68 & 1.62 & 2.15 & 22.09 & 21.70 & 24.0 \\      % 10 hexes
                & 2016-01-09.11 & 13.96 & 1.17 & 1.12 & 1.80 & 22.44 & 22.17 &  4.8 \\      %  2 hexes
                 \hline
 Epoch 4 & 2016-01-18.03 & 22.88 & 1.21 & 1.20 & 1.48 & 22.00 & 22.01 & 12.0 \\      %  5 hexes
               & 2016-01-19.01 & 23.86 & 1.29 & 1.25 & 1.71 & 21.86 & 21.90 & 16.8 \\ %  7 hexes
\enddata
\tablecomments{Summary of our DECam follow-up observations of
GW151226. The PSF, airmass, and depth are the average values across
all observations on that date. The reported depth corresponds to the mean
$5\sigma$ point source detection in the coadded search images.}
\tablenotetext{a}{Time elapsed between the GW trigger time and the time of the first image.}
\tablenotetext{b}{The effective area corresponds to 12 DECam pointings
taking into account that $\approx 20\%$ of the 3
deg$^2$ field of view of DECam is lost due to chip gaps (10\%), 3
dead CCDs (5\%, \citealt{Diehl+14}), and masked edge pixels (5\%).}
\end{deluxetable}

@@

\section{Search for an Optical Counterpart}
\label{sec:ch4_analysis}

The primary focus of our search is a fast-fading transient.
While the merger of a BBH system is not expected
to produce an EM counterpart, it is informative to consider the
possibility of optical emission due to the presence of some matter in
the system. As a generic example, we consider the behavior of a
transient such as a short gamma-ray burst (SGRB) with a typical beaming-corrected energy of
$E_j \approx 10^{49}$ erg and an opening angle of $\theta_j\approx
10^\circ$ \citep{Berger2014,Fong+15}. If viewed far off-axis $(\theta_{\rm
  obs}\gtrsim 4\theta_j)$ the optical emission will reach peak
brightness after several days, but at the distance of GW151226
($\approx440$ Mpc, \citealt{LIGOGW151226}), the peak brightness will be $i\approx 26$ mag (see Figure 5 of
\citealt{MetzgerBerger12}), well beyond our detection limit. If the source
is observed moderately off-axis or on-axis $(\theta_{\rm obs} \lesssim
2\theta_j)$, then the light curve will decline throughout our
observations, roughly as $F_\nu\propto t^{-1}$, and will be detectable
at $i\approx$ 21--22 mag in our first observation (see Figures 3 and 4
of \citealt{MetzgerBerger12}). We can apply a similar argument to the
behavior of a more isotropic (and non-relativistic) outflow given that
any material ejected in a BBH merger is likely to have a low mass and
the outflow will thus become optically thin early, leading to fading
optical emission. Based on this line of reasoning, we search our data
for steadily declining transients.

@@

Only four events passed our final criterion. We find that two of those events are
coincident with the nuclei of known AGN (PKS\,0129-066 and Mrk\,584), indicating
that they represent AGN variability. A third candidate is coincident with the nucleus of
the bright radio source PMN\,J0203+0956 ($F_\nu(365\,{\rm MHz})\approx 0.4$ Jy,
\citealt{Douglas+96}), also suggesting AGN variability.

\begin{figure}[t!]
\centering
\includegraphics[width=0.95\textwidth]{./figs/chapter4/fig2.pdf}
\caption{{\it Top}: Single-epoch images of our main candidate
from all four epochs (green circle). This is the event discovered as PS15cdi in
the PSST about 94 d prior to GW151226. {\it Bottom}: Light
curve data for PS15cdi from PSST $w$- and $i$-band
observations (green squares and yellow diamonds, respectively).
Our DECam $i$- and $z$-band data are shown as
blue circles and red stars, respectively. The revised DECam analysis
using pre-existing templates is shown as open symbols. Upper
limits are indicated by triangles. The inset focuses on our
DECam data, indicating a rapid decline in both $i$ and $z$
bands.  We fit a power-law model to the data finding a temporal index of
$\alpha = -0.43$ (dashed-dot line). Kilonova models from Barnes and
Kasen \citeyearpar{BarnesKasen13} with $v_{\rm ej}=0.2c$ and $M_{\rm ej}=0.1$ M$_\odot$
at a distance of $187$~Mpc are also shown (dashed line).}
\label{fig:ch4_PS15cdi}
\end{figure}

The final candidate in our search is located at RA = 01$^{\rm h}$42$^{\rm m}$16$^{\rm s}$.17
and DEC = $-$02\arcdeg13\arcmin42 6\arcsec~(J2000), with an offset of $5.8$ arcsec from the galaxy
CGCG 386-030 (RA =  01$^{\rm h}$42$^{\rm m}$15$^{\rm s}$.6, DEC = $-$02\arcdeg13\arcmin38
5\arcsec; J2000), at $z = 0.041$ or $d_L \approx 187$ Mpc (6dFGS, \citealt{Jones+04,Jones+09}); see
\Cref{fig:ch4_PS15cdi}. We note that this distance is inconsistent with the 90\% confidence interval for
the distance to GW151226 based on the GW data \citep{LIGOGW151226}. We observe this source in a state of rapid decline with an
absolute magnitude of $M_i \approx -15$~mag on 2015 December 28 and $M_i\approx -14.5$~mag on 2016
January 1, indicating a decline rate of $\approx 0.12$ mag d$^{-1}$; the decline rate in $z$-band is
$\approx 0.10$ mag d$^{-1}$. Additionally, the source exhibits a red $i-z$ color of $0.3$ mag. We
fit these data to a power-law model typical for GRB afterglows $(F_\nu \propto \nu^{\beta} t^{\alpha})$
and find a temporal index of $\alpha = -0.43\pm0.12$ and a spectral index of $\beta = -1.8\pm0.8$,
both of which differ from the expected values for GRB afterglows ($\alpha \approx -1$,
$\beta \approx -0.75$, \citealt{Sari+98}). Additionally, we compare our observations to a
kilonova model with ejecta parameters of $v_{\rm ej} = 0.2$c and $M_{\rm ej} = 0.1$ M$_\odot$
 \citep{BarnesKasen13}. We find that the timescale of the transient agrees with those expected for kilonovae,
 but the color is bluer than the expected value of $i-z \approx 1$~mag \citep{BarnesKasen13}. Thus, the properties
 of this transient differ from those of GRB afterglows or kilonovae. The observations
 and models are shown in \Cref{fig:ch4_PS15cdi}.

This source was previously detected as PS15cdi on 2015 September 23 by the Pan-STARRS
Survey for Transients (PSST\footnote{\url{http://psweb.mp.qub.ac.uk/ps1threepi/psdb/candidate/1014216170021342600/}},
\citealt{Huber+15}); see \Cref{fig:ch4_PS15cdi}. The absolute $i$-band
magnitude in the first PSST epoch, $M_i\approx -16.6$ mag and the shallow decline of $\approx0.6$~mag
over $\approx70$~d, are consistent with a Type IIP core-collapse supernova (SN).
A likely interpretation of the rapid decline in our observations is that PS15cdi is a Type IIP SN undergoing
the rapid transition from the hydrogen recombination driven plateau to the radioactive $^{56}$Co dominated
phase \citep{KasenWoosley09,Sanders+15,Dhungana+16}. The red $i-z$ color in our data is consistent with observations
of other IIP SN during this phase of evolution (e.g., SN2013ej, \citealt{Dhungana+16}).
This transition typically occurs about 100~d post explosion \citep{KasenWoosley09,Sanders+15,Dhungana+16}, consistent with the
timing of our observations relative to the first detection in PSST.

To mitigate the effect of excess flux from PS15cdi still present in our template observations, we repeat the analysis
using as templates archival DES $i$- and $z$-band images from 2013 December 19. These data were processed
and image subtraction was performed as described in \cref{sec:ch4_obs}. We find that flux from PS15cdi is indeed still present
in our original template image, leading to revised first epoch absolute magnitudes of $M_i \approx -15.6$ and
$M_z \approx -16$~mag, and a decline rate between the first and fourth epochs of 0.04~mag d$^{-1}$, in both
$i$- and $z$-bands. The transient still exhibits a red $i-z$ colors of $\approx 0.4$~mag across all four epochs.
